---
title: hw7 
author: Jialing Deng
date: Feb 28
output:
  pdf_document:
    toc: true
    toc_depth: 1
---

_Note:_ To help show what your homework should look like, these solutions don't
display the code used.  You should look at the Rmd file as well to see those.
(The comments to the code include some notes on why it's done the way it is, and
alternatives.)

```{r, include=FALSE}
# Standard set-up
library(knitr)
knitr::opts_chunk$set(warning=FALSE, message=FALSE)


# Set knitr options for knitting code into the report:
# - Save results so that code blocks aren't re-run unless code changes (cache),
  # _or_ a relevant earlier code block changed (autodep)
  # but don't re-run if only the comments changed (cache.comments)
# - Don't clutter R output with messages or warnings (message, warning)
  # This _will_ leave error messages showing up in the knitted report
# - By default, don't print the code (echo)
opts_chunk$set(cache=TRUE, autodep=TRUE, cache.comments=FALSE,
               message=FALSE, warning=FALSE, echo=FALSE)
```


### Q1.


### a.

Evaluating predictions only on the testing set is important because it provides 
an unbiased estimate of model performance on unseen data. If we evaluate the model 
on the same training data that was used to build the model, we risk overfitting, 
so the model may perform well on the training data but fail to generalize to 
new cases. 




### b. 

Randomly dividing the data ensures that both the training and testing sets are 
representative of the overall dataset, reducing the risk of introducing systematic 
biases. If the split is not random, certain patterns or features may be disproportionately
represented in one subset, leading to misleading performance estimates and a model 
that does not generalize well. Randomization helps create a more balanced and fair 
evaluation of the model.






```{r}
setwd("/Users/dengjialing/Downloads")

full_data <- read.csv("compas_violence.csv", header=TRUE, stringsAsFactors=FALSE)

test_indices <- scan("testing_set.txt", what=integer(), skip=1)


test_set <- full_data[test_indices, ]
train_set <- full_data[-test_indices, ]

# Save the training and testing datasets as CSV files
write.csv(train_set, "training_set.csv", row.names=FALSE)
write.csv(test_set, "testing_set_processed.csv", row.names=FALSE)

train_df<-read.csv("training_set.csv", header=TRUE, stringsAsFactors=FALSE)

test_df <- read.csv("testing_set_processed.csv", header=TRUE, stringsAsFactors=FALSE)


prop_rearrested <- mean(test_df$two_year_recid == 1)



```


### c.

18.41%  of people in the testing set are re-arrested for violence within two years. 

\newpage

### Q2.
```{r}

logit_model <- glm(two_year_recid ~ age + priors_count, data=train_df, family=binomial)
```

### a. 
```{r}

library(boot)


# Logistic regression model
logit_model <- glm(two_year_recid ~ age + priors_count, data=train_df, family=binomial)


 rboot<- function(statistic, simulator, B) {
  tboots <- replicate(B, statistic(simulator()))
  if (is.null(dim(tboots))) {
    tboots <- array(tboots, dim=c(1, B))
  }
  return(tboots)
}

bootstrap <- function(tboots, summarizer, ...) {
  summaries <- apply(tboots, 1, summarizer, ...)
  return(t(summaries))  
}

equitails <- function(x, alpha) {
  lower <- quantile(x, alpha/2)
  upper <- quantile(x, 1-alpha/2)
  return(c(lower, upper))
}

bootstrap.ci <- function(statistic=NULL, simulator=NULL, tboots=NULL,
                         B=if (!is.null(tboots)) { ncol(tboots) },
                         t.hat, level) {
  if (is.null(tboots)) {
    stopifnot(!is.null(statistic))
    stopifnot(!is.null(simulator))
    stopifnot(!is.null(B))
    tboots <- rboot(statistic, simulator, B)
  }
  
  alpha <- 1 - level
  intervals <- bootstrap(tboots, summarizer=equitails, alpha=alpha)
  
  upper <- t.hat + (t.hat - intervals[,1])
  lower <- t.hat + (t.hat - intervals[,2])
  
  return(cbind(lower=lower, upper=upper))
}


logistic_estimator <- function(data) {
  model <- glm(two_year_recid ~ age + priors_count, data=data, family=binomial)
  return(coef(model))  
}


resample.cases <- function() {
  resampled_data <- train_df[sample(1:nrow(train_df), replace=TRUE), ]
  return(resampled_data)
}

# Bootstrap confidence intervals for logistic regression coefficients
set.seed(123) 
logistic_ci <- bootstrap.ci(
  statistic = logistic_estimator,
  simulator = resample.cases,
  B = 400,
  t.hat = logistic_estimator(train_df),
  level = 0.95
)


coef_table <- data.frame(
  Coefficient = names(coef(logit_model)),
  Estimate = coef(logit_model),
  Lower_95_CI = logistic_ci[, "lower"],
  Upper_95_CI = logistic_ci[, "upper"]
)






```
### Regression Coefficients

The estimated regression coefficients along with their 95% confidence intervals are:

Intercept (\(\beta_0\)): 
  \[
  \hat{\beta_0} = -0.156, \quad 95\% \text{ CI: } [-0.5482, 0.2370]
  \]

Age (\(\beta_1\)):    \[
  \hat{\beta_1} = -0.061, \quad 95\% \text{ CI: } [-0.0728, -0.0465]
  \]

Priors Count(\(\beta_2\)):  
  \[
  \hat{\beta_2} = 0.162, \quad 95\% \text{ CI: } [0.1381, 0.1825]
  \]






### b.


- \( \log \left( \frac{P(Y=1)}{P(Y=0)} \right) \) represents the log-odds of recidivism.
- \( \beta_0 = -0.155 \) (Intercept) is the baseline log-odds of recidivism when 
both age and priors are zero.
- \( \beta_1 = -0.061 \) means that for every additional year of age, the l
og-odds of recidivism decrease by 0.061. This suggests that older individuals 
are less likely to recidivate.
- \( \beta_2 = 0.162 \) means that each additional prior conviction increases 
the log-odds of recidivism by 0.162. This suggests that having more prior offenses
leads to a higher chance of re-arrest.



### c.
```{r}
library(mgcv)  
library(boot)  


new_data <- data.frame(
  age = c(19, 22, 34, 42),
  priors_count = c(1, 2, 0, 12)  
)


new_data <- data.frame(
  age = c(19, 22, 34, 42),
  priors_count = c(1, 2, 0, 12) )


pred_probs <- predict(logit_model, new_data, type="response")


logit_estimator <- function(data) {
  model <- glm(two_year_recid ~ age + priors_count, data=data, family=binomial)
  predict(model, new_data, type="response")  
}


resample.cases <- function() {
  resampled_data <- train_df[sample(1:nrow(train_df), replace=TRUE), ]
  return(resampled_data)
}


set.seed(123)
boot_preds <- rboot(statistic=logit_estimator, simulator=resample.cases, B=400)


ci_table <- bootstrap(tboots=boot_preds, summarizer=equitails, alpha=0.05)


results <- data.frame(
  Individual = c("Archie", "Betty", "Chuck", "Veronica"),
  Predicted_Probability = pred_probs,
  CI_Lower = ci_table[, 1],
  CI_Upper = ci_table[, 2]
)



```



For Archie, \( P(\text{Recidivism}) = 0.241 \) with confidence interval \([0.211, 0.273]\).

For Betty, \( P(\text{Recidivism}) = 0.237 \) with confidence interval \([0.212, 0.263]\).

For Chuck, \( P(\text{Recidivism}) = 0.098 \) with confidence interval \([0.085, 0.109]\).

For Veronica, \( P(\text{Recidivism}) = 0.317 \) with confidence interval \([0.272, 0.363]\).

\newpage

### Q3. 

### a. 

```{r}

library(dbplyr)
library(tidyverse)
library(ggplot2)



test_probs <- predict(logit_model, test_df, type="response")


evaluate_threshold <- function(t) {
  predictions <- ifelse(test_probs >= t, 1, 0)  
  actual <- as.numeric(test_df$two_year_recid)  
  correct <- sum(predictions == actual) / length(actual)  
  FP <- sum(predictions == 1 & actual == 0)
  TN <- sum(predictions == 0 & actual == 0)
  FPR <- FP / (FP + TN)

  FN <- sum(predictions == 0 & actual == 1)
  TP <- sum(predictions == 1 & actual == 1)
  FNR <- FN / (FN + TP)

  return(c(t, correct, FPR, FNR))
}


thresholds <- seq(0, 1, length.out=101)
eval_results <- t(sapply(thresholds, evaluate_threshold))


colnames(eval_results) <- c("Threshold", "Accuracy", "FPR", "FNR")
eval_results <- as.data.frame(eval_results)


best_threshold <- eval_results$Threshold[which.max(eval_results$Accuracy)]
best_accuracy <- max(eval_results$Accuracy)


baseline_accuracy <- max(mean(test_df$two_year_recid == "1"), mean(test_df$two_year_recid == "0"))




```

### a.

```{r}
ggplot(eval_results, aes(x=Threshold, y=Accuracy)) +
  geom_line(color="blue") +
  geom_point(color="black") +  # Add dots at each threshold
  geom_hline(yintercept=baseline_accuracy, linetype="dashed", color="red") +
  labs(title="Accuracy vs. Threshold for Logistic", x="Threshold", y="Accuracy") +
  annotate("text", x=0.5, y=baseline_accuracy + 0.02, label="Baseline Accuracy", color="red") +
  theme_minimal()

```

At threshold = 0.39 is the accuracy maximized. The horizontal line should return 1. 
The baseline accuracy is 0.816. 



### b.

```{r}


ggplot(eval_results, aes(x=Threshold, y=FPR)) +
  geom_line(color="red") +
  geom_point(color="black") +  # Add dots at each threshold
  labs(title="False Positive Rate vs. Threshold", x="Threshold", y="FPR") +
  theme_minimal()






```
FPR is minimized when threshold is equal to 1. 


### c. 
```{r}

ggplot(eval_results, aes(x=Threshold, y=FNR)) +
  geom_line(color="green") +
  geom_point(color="black") +  # Add dots at each threshold
  labs(title="False Negative Rate vs. Threshold", x="Threshold", y="FNR") +
  theme_minimal()


```
FNR is minimized when threshold is equal to 0. 

### d. 

```{r}
ggplot(eval_results, aes(x=FPR, y=FNR)) +
  geom_line(color="purple") +
  geom_point(color="black") +  # Add dots at each point
  geom_abline(slope=-1, intercept=1, linetype="dashed", color="gray") +
  labs(title="FNR vs. FPR", x="False Positive Rate (FPR)", y="False Negative Rate (FNR)") +
  annotate("text", x=0.5, y=0.5, label="Random Classifier Line", color="gray") +
  theme_minimal()
```

The curve stays below the dashed line, indicating that logistic regression
provides better predictive power than random guessing.
The model reduces FNR more effectively for the same level of FPR, meaning it 
identifies more true positives while keeping false positives manageable.










\newpage





### Q4. 


### a.
```{r}

library(mgcv)

gam_model <- gam(two_year_recid ~ s(age) + s(priors_count), family=binomial, data=train_df)



plot(gam_model,se=TRUE,shade=TRUE)

```

### b.

As age increases, the risk of recidivism gradually declines.
The decline is relatively steady until about 60 years old, beyond which the effect
is less certain (wider confidence intervals).
The curve flattens slightly for middle-aged individuals, suggesting a slower decline in 
recidivism risk for this group.


The function shows a positive relationship between the number of prior convictions 
and predicted recidivism risk.
Initially, as priors_count increases from 0 to around 10, the recidivism risk 
increases steadily. Beyond 10 priors, the relationship flattens somewhat, suggesting
that having more priors still increases recidivism risk but at a slower rate.



### c.


```{r}
library(mgcv)  
library(boot)  


new_data <- data.frame(
  age = c(19, 22, 34, 42),
  priors_count = c(1, 2, 0, 12)  
)


gam_model <- gam(two_year_recid ~ s(age) + s(priors_count), data=train_df, family=binomial)


new_data <- data.frame(
  age = c(19, 22, 34, 42),
  priors_count = c(1, 2, 0, 12)  
)


pred_probs <- predict(gam_model, new_data, type="response")


gam_estimator <- function(data) {
  model <- gam(two_year_recid ~ s(age) + s(priors_count), data=data, family=binomial)
  predict(model, new_data, type="response")  
}


resample.cases <- function() {
  resampled_data <- train_df[sample(1:nrow(train_df), replace=TRUE), ]
  return(resampled_data)
}


set.seed(123)
boot_preds <- rboot(statistic=gam_estimator, simulator=resample.cases, B=400)


ci_table <- bootstrap(tboots=boot_preds, summarizer=equitails, alpha=0.05)


results <- data.frame(
  Individual = c("Archie", "Betty", "Chuck", "Veronica"),
  Predicted_Probability = pred_probs,
  CI_Lower = ci_table[, 1],
  CI_Upper = ci_table[, 2]
)



```

For Archie, the predicted probability of recidivism is:
\[
P(\text{Recidivism}) = 0.378
\]
with a 95% confidence interval of:
\[
[0.236, 0.563]
\]

For Betty, the predicted probability of recidivism is:
\[
P(\text{Recidivism}) = 0.292
\]
with a 95% confidence interval of:
\[
[0.249, 0.362]
\]

For Chuck, the predicted probability of recidivism is:
\[
P(\text{Recidivism}) = 0.073
\]
with a 95% confidence interval of:
\[
[0.052, 0.092]
\]

For Veronica, the predicted probability of recidivism is:
\[
P(\text{Recidivism}) = 0.298
\]
with a 95% confidence interval of:
\[
[0.203, 0.415]
\]


\newpage


### Q5.


```{r}

gam_probs <- predict(gam_model, test_df, type="response")

evaluate_model <- function(pred_probs) {
  thresholds <- seq(0, 1, length.out=101)
  evaluate_threshold <- function(t) {
    predictions <- ifelse(pred_probs >= t, 1, 0) 
    actual <- as.numeric(test_df$two_year_recid)  
    
 
    correct <- sum(predictions == actual) / length(actual) 
    
  
    FP <- sum(predictions == 1 & actual == 0)
    TN <- sum(predictions == 0 & actual == 0)
    FPR <- FP / (FP + TN)
    
    
    FN <- sum(predictions == 0 & actual == 1)
    TP <- sum(predictions == 1 & actual == 1)
    FNR <- FN / (FN + TP)
    
    return(c(t, correct, FPR, FNR))
  }
  
  
  eval_results <- t(sapply(thresholds, evaluate_threshold))
  colnames(eval_results) <- c("Threshold", "Accuracy", "FPR", "FNR")
  
  return(as.data.frame(eval_results))
}


gam_eval_results <- evaluate_model(gam_probs)


best_gam_threshold <- gam_eval_results$Threshold[which.max(gam_eval_results$Accuracy)]
best_gam_accuracy <- max(gam_eval_results$Accuracy)



baseline_accuracy <- max(mean(test_df$two_year_recid == "1"), mean(test_df$two_year_recid == "0"))


```


### a. 
```{r}
ggplot(gam_eval_results, aes(x=Threshold, y=Accuracy)) +
  geom_line(color="blue") +
  geom_point(color="black") +  # Add dots at each threshold
  geom_hline(yintercept=baseline_accuracy, linetype="dashed", color="red") +
  labs(title="GAM: Accuracy vs. Threshold", x="Threshold", y="Accuracy") +
  annotate("text", x=0.5, y=baseline_accuracy + 0.02, label="Baseline Accuracy", color="red") +
  theme_minimal()

```

\[
\text{At threshold } = 0.45, \text{ the accuracy is maximized at } 83.58\%.
\]



### b. 
```{r}
ggplot(gam_eval_results, aes(x=FPR, y=FNR)) +
  geom_line(color="purple") +
  geom_point(color="black") +  # Add dots at each point
  geom_abline(slope=-1, intercept=1, linetype="dashed", color="gray") +
  labs(title="GAM: FNR vs. FPR", x="False Positive Rate (FPR)", y="False Negative Rate (FNR)") +
  annotate("text", x=0.5, y=0.5, label="Random Classifier", color="gray") +
  theme_minimal()

```

### c. 
```{r}

logit_probs <- predict(logit_model, test_df, type="response")
logit_eval_results <- evaluate_model(logit_probs)


ggplot() +
  geom_line(data=logit_eval_results, aes(x=Threshold, y=Accuracy), color="blue") +
  geom_line(data=gam_eval_results, aes(x=Threshold, y=Accuracy), color="green") +
  geom_point(data=logit_eval_results, aes(x=Threshold, y=Accuracy), color="black") +
  geom_point(data=gam_eval_results, aes(x=Threshold, y=Accuracy), color="black") +
  geom_hline(yintercept=baseline_accuracy, linetype="dashed", color="red") +
  labs(title="Comparison: GAM vs Logistic Regression (Accuracy)", x="Threshold", y="Accuracy") +
  theme_minimal() +
  annotate("text", x=0.3, y=max(logit_eval_results$Accuracy), label="Logistic Regression", color="blue") +
  annotate("text", x=0.7, y=max(gam_eval_results$Accuracy), label="GAM", color="green")

```
At low thresholds (t < 0.2), GAM (green) outperforms logistic regression (blue)  
slightly. This suggests that GAM provides better accuracy when predicting a higher 
number of recidivists. At moderate thresholds (0.2 to 0.5), logistic regression and
GAM perform very  similarly, with logistic regression slightly higher in some regions. 

At high thresholds (t > 0.5), both models converge to the same accuracy.  
This indicates that when fewer individuals are classified as recidivists, both 
models make similar predictions. 

\newpage

### Q6.


```{r}

library(tree)



train_df$two_year_recid <- as.factor(train_df$two_year_recid)


tree_model <- tree(as.factor(two_year_recid) ~. -v_score, data=train_df)

```
### a. 
```{r}

plot(tree_model)
text(tree_model, pretty=0, cex=0.8) 
```

### b.

The tree model used prior counts and age to make predictions. 


### c. 

If an individual has fewer than 2.5 prior convictions, the model checks their age. 
If an individual has 2.5 or more prior convictions, they are predicted as not recidivating. 
Among individuals with fewer than 2.5 priors, the model checks if they are 
younger than  21.5 years old. If age is above 21.5 but below 38.5, the model makes 
another check. However, all individuals still end up classified as non-recidivists.
Among  individuals with more than 2.5 priors, the model checks if they are under 36.5 
years old. However, all cases result in a classification of 0 (no recidivism).  
This tree is trivial and ineffective because it predicts all individuals as 
non-recidivists (0). The tree suggests that age and priors_count alone do not
provide enough discriminatory power to separate recidivists from non-recidivists.  







### d. 

```{r,warning=FALSE}
library(tree)


train_df$two_year_recid <- as.factor(train_df$two_year_recid)

new_data_tree <- data.frame(
  age = c(19, 22, 34, 42),
  priors_count = c(1, 2, 0, 12),
  charge_degree = factor(c("F", "M", "M", "F")), 
  sex = factor(c("Male", "Female", "Male", "Female")),
  race = factor(c("Caucasian", "Caucasian", "African-American", "Hispanic"))
)

tree_df <- train_df[, !(names(train_df) %in% c("v_score"))]

tree_model <- tree((two_year_recid) ~., data=tree_df)


 rboot<- function(statistic, simulator, B) {
  tboots <- replicate(B, statistic(simulator()))
  if (is.null(dim(tboots))) {
    tboots <- array(tboots, dim=c(1, B))
  }
  return(tboots)
}

tree_estimator <- function(data=new_data_tree) {
  tree_model<-tree(two_year_recid ~ ., data=data)
  predict(tree_model, new_data_tree, family="binomial") [,2] 
}


resample.cases <- function() {
  resampled_data <- tree_df[sample(1:nrow(tree_df), replace=TRUE), ]
  return(resampled_data)
}


set.seed(402)
boot_preds <- rboot(statistic=tree_estimator, simulator=resample.cases, B=400)


ci_table <- bootstrap(tboots=boot_preds, summarizer=equitails, alpha=0.05)


results <- data.frame(
  Individual = c("Archie", "Betty", "Chuck", "Veronica"),
  Predicted_Probability = pred_probs,
  CI_Lower = ci_table[, 1],
  CI_Upper = ci_table[, 2]
)


```

For Archie, the predicted probability of recidivism is:
\[
P(\text{Recidivism}) = 0.378
\]
with a 95% confidence interval of:
\[
[0.12, 1.00]
\]

For Betty, the predicted probability of recidivism is:
\[
P(\text{Recidivism}) = 0.292
\]
with a 95% confidence interval of:
\[
[0.08, 0.34]
\]

For Chuck, the predicted probability of recidivism is:
\[
P(\text{Recidivism}) = 0.073
\]
with a 95% confidence interval of:
\[
[0.05, 0.15]
\]

For Veronica, the predicted probability of recidivism is:
\[
P(\text{Recidivism}) = 0.298
\]
with a 95% confidence interval of:
\[
[0.12, 0.50]
\]



### Q7.
```{r}
library(ggplot2)


tree_probs <- predict(tree_model, test_df, family="binomial")[,2]


evaluate_model <- function(pred_probs, actual_values) {
  thresholds <- seq(0, 1, length.out=101)  
  
  evaluate_threshold <- function(t) {
    predictions <- ifelse(pred_probs >= t, 1, 0)  
    actual <- as.numeric(actual_values)  

    
    correct <- sum(predictions == actual) / length(actual)

    
    FP <- sum(predictions == 1 & actual == 0)
    TN <- sum(predictions == 0 & actual == 0)
    FPR <- ifelse((FP + TN) > 0, FP / (FP + TN), 0)

    
    FN <- sum(predictions == 0 & actual == 1)
    TP <- sum(predictions == 1 & actual == 1)
    FNR <- ifelse((FN + TP) > 0, FN / (FN + TP), 0)

    return(c(t, correct, FPR, FNR))
  }

 
  eval_results <- t(sapply(thresholds, evaluate_threshold))
  colnames(eval_results) <- c("Threshold", "Accuracy", "FPR", "FNR")

  return(as.data.frame(eval_results))
}


tree_eval_results <- evaluate_model(tree_probs, test_df$two_year_recid)


best_tree_threshold <- tree_eval_results$Threshold[which.max(tree_eval_results$Accuracy)]
best_tree_accuracy <- max(tree_eval_results$Accuracy)


baseline_accuracy <- max(mean(test_df$two_year_recid == "1"), mean(test_df$two_year_recid == "0"))


```


### a. 
```{r}

ggplot(tree_eval_results, aes(x=Threshold, y=Accuracy)) +
  geom_point(color="black") + geom_line(color="blue") +
  geom_hline(yintercept=baseline_accuracy, linetype="dashed", color="red", linewidth=1) +
  labs(title="Decision Tree: Accuracy vs Threshold", 
       x="Threshold", 
       y="Accuracy") +
  theme_minimal() +
  annotate("text", x=0.5, y=baseline_accuracy + 0.02, 
           label=sprintf("Baseline Accuracy = %.2f", baseline_accuracy), color="red")


```

### b. 

```{r}
ggplot(tree_eval_results, aes(x=FPR, y=FNR)) +
  geom_point(color="black") + geom_line(color="blue") +
  geom_abline(slope=-1, intercept=1, linetype="dashed", color="red") +
  labs(title="Decision Tree: FPR vs FNR", 
       x="False Positive Rate", 
       y="False Negative Rate") +
  theme_minimal()

```


The plot for the Decision Tree Model has fewer than 101 points because decision 
trees make discrete predictions, meaning that the probability outputs are not 
continuous like in logistic regression or GAM models.






### c. 

The Decision Tree model has significantly fewer threshold points, leading to a 
less smooth FNR vs. FPR curve. The Logistic Regression and GAM models produce a 
much smoother FNR vs. FPR curve with more granularity. The GAM model appears to 
provide more flexibility across different thresholds.


For the decision tree, FPR and FNR do not get very close to zero simultaneously.
This suggests that the tree model struggles with a balanced trade-off. 

The GAM model achieves lower FNR at a given FPR compared to the Decision Tree.
At high thresholds, both FPR and FNR get much closer to zero compared to the Decision Tree.

The logistic regression model behaves similarly to GAM but appears slightly worse off. 




\newpage

### Q8.
```{r}
library(ggplot2)

compas_scores <- test_df$v_score  
actual_values <- as.numeric(test_df$two_year_recid)  

compas_thresholds <- 1:11  


compute_accuracy <- function(t) {
  predictions <- ifelse(compas_scores >= t, 1, 0)
  mean(predictions == actual_values)
}

compas_accuracies <- sapply(compas_thresholds, compute_accuracy)

compas_accuracy_df <- data.frame(
  Threshold = compas_thresholds,
  Accuracy = compas_accuracies
)
```


```{r}
ggplot(compas_accuracy_df, aes(x=Threshold, y=Accuracy)) +
  geom_point(color="black") + geom_line(color="blue") +
  geom_hline(yintercept=max(compas_accuracies), linetype="dashed", color="red") +
  labs(title="COMPAS: Accuracy vs Threshold", x="Threshold (t)", y="Accuracy") +
  theme_minimal()

```

When we set a threshold t, we predict that someone will be arrested for a violent 
crime only if their COMPAS score is at least t. At t = 11, since no one has a 
COMPAS score of 11, we predict that no one will be arrested for violence.
Setting t=11 ensures that we see the full transition from always predicting 
recidivism (t = 1) to never predicting recidivism (t = 11). 



### b. 
```{r}

compute_fpr_fnr <- function(t) {
  predictions <- ifelse(compas_scores >= t, 1, 0)
  
  fp <- sum(predictions == 1 & actual_values == 0)
  tn <- sum(predictions == 0 & actual_values == 0)
  fpr <- ifelse((fp + tn) > 0, fp / (fp + tn), 0)

  fn <- sum(predictions == 0 & actual_values == 1)
  tp <- sum(predictions == 1 & actual_values == 1)
  fnr <- ifelse((fn + tp) > 0, fn / (fn + tp), 0)

  return(c(fpr, fnr))
}


compas_fpr_fnr_values <- t(sapply(compas_thresholds, compute_fpr_fnr))


compas_fpr_fnr_df <- data.frame(
  FPR = compas_fpr_fnr_values[,1],
  FNR = compas_fpr_fnr_values[,2]
)


ggplot(compas_fpr_fnr_df, aes(x=FPR, y=FNR)) +
  geom_point(color="black") + geom_line(color="blue") +
  geom_abline(slope=-1, intercept=1, linetype="dashed", color="red") +
  labs(title="COMPAS: FPR vs FNR", x="False Positive Rate", y="False Negative Rate") +
  theme_minimal()

```


### c.

The dashed red line shows the baseline accuracy, meaning COMPAS does not 
significantly outperform it. However, GAM peaks above the baseline accuracy. The
GLM also peaks near the baseline accuracy. On the other hand, the accuracy for 
decision tree fluctuates significantly, so it is not as consistent as COMPAS or GAM
or logistic model. In general, GAM achieves high accuracy more efficiently than COMPAS.
COMPAS is more stable than the Decision Tree but does not outperform GAM and GLM.


\newpage

### Q9. 

```{r}


logit_model <- glm(two_year_recid ~ age + priors_count, data=test_df)
frequency.vs.probability <- function(p.lower, p.increment,
  model, events) {
  fitted.probs <- fitted(model)


  indices <- (fitted.probs >= p.lower) & (fitted.probs < p.lower+p.increment)
 

  ave.prob <- mean(fitted.probs[indices])
  frequency <- mean(events[indices])
  individual.vars <- fitted.probs[indices]*(1-fitted.probs[indices])
  var.of.average <- sum(individual.vars)/(sum(indices)^2)
  se <- sqrt(var.of.average)
  out <- c(frequency=frequency, ave.prob=ave.prob, se=se)
  return(out)
}


```




```{r}



f.vs.p <- sapply(seq(0, 8/10, by = 0.1), frequency.vs.probability, p.increment=0.1,
                 model=logit_model, events=test_df$two_year_recid)
f.vs.p <- data.frame(frequency=f.vs.p["frequency",],
                     ave.prob=f.vs.p["ave.prob",],
                     se=f.vs.p["se",])

```

### a. 
```{r}
par(pty="s") # Square plotting window
plot(frequency~ave.prob, data=f.vs.p, xlab="Predicted probabilities",
     ylab="Observed frequencies", xlim=c(0,1), ylim=c(0,1))
points(f.vs.p$ave.prob, f.vs.p$frequency, pch=16, col="blue")
rug(fitted(logit_model), col="lightgrey")
abline(0,1, col="darkgrey")
segments(x0=f.vs.p$ave.prob, y0=f.vs.p$ave.prob-1.96*f.vs.p$se,
                             y1=f.vs.p$ave.prob+1.96*f.vs.p$se)
```



```{r}


gam_model<-gam(two_year_recid ~ s(age) + s(priors_count), family=binomial, data=test_df)
f.vs.p <- sapply(0:8/10, frequency.vs.probability, p.increment=0.1,
                 model=gam_model, events=test_df$two_year_recid)
f.vs.p <- data.frame(frequency=f.vs.p["frequency",],
                     ave.prob=f.vs.p["ave.prob",],
                     se=f.vs.p["se",])

par(pty="s")
plot(frequency~ave.prob, data=f.vs.p, xlab="Predicted probabilities",
     ylab="Observed frequencies", xlim=c(0,1), ylim=c(0,1))

points(f.vs.p$ave.prob, f.vs.p$frequency, pch=16, col="blue")
rug(fitted(gam_model),col="lightgrey")
abline(0,1,col="darkgrey")
segments(x0=f.vs.p$ave.prob,y0=f.vs.p$ave.prob-1.96*f.vs.p$se,
  y1=f.vs.p$ave.prob+1.96*f.vs.p$se)
```


```{r}
library(tree)

frequency.vs.probability <- function(p.lower, p.increment, model, events) {
  fitted.probs <- predict(model, type="vector")[,2]  

  indices <- (fitted.probs >= p.lower) & (fitted.probs < p.lower + p.increment)
   ave.prob <- mean(fitted.probs[indices], na.rm = TRUE)
  frequency <- mean(events[indices], na.rm = TRUE)
  individual.vars <- fitted.probs[indices] * (1 - fitted.probs[indices])
  var.of.average <- sum(individual.vars, na.rm = TRUE) / (sum(indices, na.rm = TRUE)^2)
  se <- sqrt(var.of.average)

  return(c(frequency=frequency, ave.prob=ave.prob, se=se))
}



tree_model <- tree(two_year_recid ~., data=tree_df)

fitted.probs <- predict(tree_model, test_df, family="binomial")[,2]
 



f.vs.p <- sapply(seq(0, 0.6, by = 0.1), frequency.vs.probability, p.increment=0.1,
                 model=tree_model, events=test_df$two_year_recid)

f.vs.p <- data.frame(frequency=f.vs.p["frequency",],
                     ave.prob=f.vs.p["ave.prob",],
                     se=f.vs.p["se",])

par(pty="s") # Square plotting window
plot(frequency~ave.prob, data=f.vs.p, xlab="Predicted probabilities",
     ylab="Observed frequencies", xlim=c(0,1), ylim=c(0,1))
points(f.vs.p$ave.prob, f.vs.p$frequency, pch=16, col="blue")
rug(fitted(logit_model), col="lightgrey")
abline(0,1, col="darkgrey")
segments(x0=f.vs.p$ave.prob, y0=f.vs.p$ave.prob-1.96*f.vs.p$se,
                             y1=f.vs.p$ave.prob+1.96*f.vs.p$se)
```

### b.

The GAM model appears to be the best-calibrated overall, as it follows the diagonal 
line closely. The logistic regression model is slightly worse, especially at high 
probabilities (0.8). 


Comparing the error bars, the decision tree has the least amount of error, but 
there are too few points, so hard to make generalized conclusion.The error bar 
for logistic model and the gam are both larger than the decision tree, while the 
gam has slightly smaller error than the logistic model. In conclusion, the GAM is 
more well calibrated, the decision tree is also well calibrated by looking at the only 
few predictions. The logistic model is worse calibrated than the other two models. 





### c.


```{r}
library(tidyverse)
compas_summary <- test_df %>%
  group_by(v_score) %>%
  summarise(
    frequency = mean(two_year_recid),
    n = n(),
    se = sqrt(frequency * (1 - frequency) / n)
  )


ggplot(compas_summary, aes(x=v_score, y=frequency)) +
  geom_point(color="blue") +
  geom_errorbar(aes(ymin=frequency - 2*se, ymax=frequency + 2*se), width=0.2) +
  labs(x="COMPAS Score", y="Recidivism Frequency",
       title="Observed Recidivism vs. COMPAS Score") +
  theme_minimal()

```


The standard error (SE) of the observed recidivism frequency is calculated using the formula:

$$ SE = \sqrt{\frac{p(1 - p)}{n}} $$

where:

\( p \) is the observed recidivism frequency for each COMPAS score,
 \( n \) is the number of individuals with that specific score.

 Since the variance of a binomial proportion is given by \( p(1 - p) \), 
 dividing by the sample size \( n \) gives the estimated variance of the sample mean. 
 



### d.

The COMPAS score is an ordinal risk score, not a probability, so they are not 
modeled as calibrated probabilities between 0 and 1. Calibration plots typically 
check whether the model outputs reliable probability estimates, whereas the 
COMPAS plot merely shows an empirical relationship between a score and an outcome.





### e.

The COMPAS score is designed to reflect the risk of recidivism, with higher 
scores indicating a greater likelihood of reoffending.The general trend from the plot
appears to be increasing, meaning that individuals with higher COMPAS scores tend 
to have a higher observed recidivism rate. However, there are irregularities at some points 
(e.g., Score 10 appears to drop slightly).
The error bars  indicate uncertainty. 


\newpage

### Q10.


### a.



We start with the conditional expectation of \( R \) given \( X = x \):

\[
\mathbb{E}[R | X = x] = \mathbb{E} \left[ \frac{Y - \mu(X)}{\sqrt{\mu(X)(1 - \mu(X))}} \Bigg| X = x \right]
\]



By definition, the conditional expectation \( \mu(X) \) is given by:

\[
\mu(X) = \mathbb{E}[Y | X = x]
\]

Substituting this into the equation:

\[
\mathbb{E}[R | X = x] = \mathbb{E} \left[ \frac{Y - \mathbb{E}[Y | X = x]}{\sqrt{\mu(X)(1 - \mu(X))}} \Bigg| X = x \right]
\]


Using the linearity property of expectation:

\[
\mathbb{E} \left[ \frac{Y - \mathbb{E}[Y | X = x]}{\sqrt{\mu(X)(1 - \mu(X))}} \Bigg| X = x \right]
= \frac{1}{\sqrt{\mu(X)(1 - \mu(X))}} \mathbb{E} \left[ Y - \mathbb{E}[Y | X = x] \Bigg| X = x \right]
\]


By definition of conditional expectation:

\[
\mathbb{E}[Y | X = x] = \mu(X)
\]

Since \( \mathbb{E}[Y | X = x] \) is a deterministic quantity (a function of \( x \)), we can factor it out:

\[
\mathbb{E} \left[ Y - \mathbb{E}[Y | X = x] \Bigg| X = x \right]
= \mathbb{E}[Y | X = x] - \mathbb{E}[Y | X = x] = 0
\]



Substituting this result back:

\[
\mathbb{E}[R | X = x] = \frac{0}{\sqrt{\mu(X)(1 - \mu(X))}} = 0
\]

Thus, we conclude:

\[
\mathbb{E}[R | X = x] = 0.
\]

This proves that the expected value of the normalized residual \( R \) is zero given \( X = x \).


### Variance. 
Using the variance definition:

\[
\text{Var}[R | X = x] = \mathbb{E}[R^2 | X = x] - (\mathbb{E}[R | X = x])^2.
\]

Since we already proved \( \mathbb{E}[R | X = x] = 0 \), the variance simplifies to:

\[
\text{Var}[R | X = x] = \mathbb{E}[R^2 | X = x].
\]

Now, squaring the residual definition:

\[
R^2 = \frac{(Y - \mu(X))^2}{\mu(X)(1 - \mu(X))}.
\]

Taking expectation:

\[
\mathbb{E}[R^2 | X = x] = \mathbb{E} \left[ \frac{(Y - \mu(X))^2}{\mu(X)(1 - \mu(X))} \Bigg| X = x \right].
\]

Since \( \frac{1}{\mu(X)(1 - \mu(X))} \) is a constant when conditioning on \( X = x \), we factor it out:

\[
\mathbb{E}[R^2 | X = x] = \frac{1}{\mu(X)(1 - \mu(X))} \mathbb{E}[(Y - \mu(X))^2 | X = x].
\]

By the definition of conditional variance,

\[
\mathbb{E}[(Y - \mu(X))^2 | X = x] = \text{Var}[Y | X = x].
\]

Since \( Y \) follows a Bernoulli distribution with probability \( \mu(X) \), we use the Bernoulli variance formula:

\[
\text{Var}[Y | X = x] = \mu(X)(1 - \mu(X)).
\]

Thus,

\[
\mathbb{E}[R^2 | X = x] = \frac{\mu(X)(1 - \mu(X))}{\mu(X)(1 - \mu(X))} = 1.
\]

Therefore, we have shown:

\[
\text{Var}[R | X = x] = 1.
\]





## Question b


The residual is defined as:

\[
R = \frac{Y - \hat{p}(X)}{\sqrt{\hat{p}(X)(1 - \hat{p}(X))}}
\]

where \( \hat{p}(X) \) is the predicted probability of \( Y = 1 \) given \( X \).


The expected value:

\[
\mathbb{E}[R | X] = \frac{\mathbb{E}[Y | X] - \hat{p}(X)}{\sqrt{\hat{p}(X)(1 - \hat{p}(X))}} = \frac{\hat{p}(X) - \hat{p}(X)}{\sqrt{\hat{p}(X)(1 - \hat{p}(X))}} = 0.
\]

The variance:

\[
\text{Var}[R | X] = \frac{\text{Var}[Y | X]}{\hat{p}(X)(1 - \hat{p}(X))} = 
\frac{\hat{p}(X)(1 - \hat{p}(X))}{\hat{p}(X)(1 - \hat{p}(X))} = 1.
\]

Thus, the residual  always has mean 0 and variance 1 when the model is correct.






### c.

```{r}
logit_model <- glm(two_year_recid ~ age + priors_count, data=train_df, family=binomial)
tree_model <- tree(as.factor(two_year_recid) ~ . -v_score, data=train_df)
gam_model<-gam(two_year_recid ~ s(age) + s(priors_count), family=binomial, data=train_df)


logit_preds <- predict(logit_model, newdata = test_df, type = "response")
logit_residuals <- (test_df$two_year_recid - logit_preds) / sqrt(logit_preds * (1 - logit_preds))


gam_preds <- predict(gam_model, newdata = test_df, type = "response")
gam_residuals <- (test_df$two_year_recid - gam_preds) / sqrt(gam_preds * (1 - gam_preds))


tree_preds <- predict(tree_model, newdata = test_df, type = "vector")[, 2]  
tree_residuals <- (test_df$two_year_recid - tree_preds) / sqrt(tree_preds * (1 - tree_preds))


residuals_df <- data.frame(
  Probability = c(logit_preds, gam_preds, tree_preds),
  Residuals = c(logit_residuals, gam_residuals, tree_residuals),
  Model = rep(c("Logistic Regression", "GAM", "Decision Tree"), each = nrow(test_df))
)

```
```{r}


plot(logit_preds, logit_residuals, col = "red", pch = 16, cex = 0.6,
     xlab = "Predicted Probability", ylab = "Residuals",
     main = "Residuals vs. Predicted Probabilities (Logistic Regression)")
lines(smooth.spline(logit_preds, logit_residuals), col = "darkred", lwd = 2)
abline(h = 0, lty = 2, col = "gray")


plot(gam_preds, gam_residuals, col = "skyblue", pch = 16, cex = 0.6,
     xlab = "Predicted Probability", ylab = "Residuals",
     main = "Residuals vs. Predicted Probabilities (GAM)")
lines(smooth.spline(gam_preds, gam_residuals), col = "blue", lwd = 2)
abline(h = 0, lty = 2, col = "gray")


plot(tree_preds, tree_residuals, col = "black", pch = 16, cex = 0.6,
     xlab = "Predicted Probability", ylab = "Residuals",
     main = "Residuals vs. Predicted Probabilities (Decision Tree)")
lines(smooth.spline(tree_preds, tree_residuals), col = "black", lwd = 2)
abline(h = 0, lty = 2, col = "gray")

```

Among the three models, the decision tree model has residuals that are the closest 
to being flat around 0, suggesting that it might be making relatively unbiased 
predictions. However, its limited range of predicted probabilities may indicate 
underfitting. The logistic regression and GAM models show systematic patterns in 
their residuals, implying some level of model bias or mis-specification.


### d.
```{r}

logit_preds <- predict(logit_model, newdata = test_df, type = "response")
logit_residuals <- (test_df$two_year_recid - logit_preds) / sqrt(logit_preds * (1 - logit_preds))


gam_preds <- predict(gam_model, newdata = test_df, type = "response")
gam_residuals <- (test_df$two_year_recid - gam_preds) / sqrt(gam_preds * (1 - gam_preds))


tree_preds <- predict(tree_model, newdata = test_df, type = "vector")[, 2]  
tree_residuals <- (test_df$two_year_recid - tree_preds) / sqrt(tree_preds * (1 - tree_preds))
test_df$logit_residuals_sq <- logit_residuals^2
test_df$gam_residuals_sq <- gam_residuals ^2
test_df$tree_residuals_sq <- tree_residuals ^2


test_df$logit_preds <- predict(logit_model, test_df, type="response")
test_df$gam_preds <- predict(gam_model, test_df, type="response")
test_df$tree_preds <- predict(tree_model, test_df, type="vector")[,2]

plot_residuals_base <- function(preds, residuals, model_name, color) {
  

  plot(preds, residuals, 
       main = paste("Squared Residuals vs. Predicted Probabilities (", model_name, ")", sep=""),
       xlab = "Predicted Probability", 
       ylab = "Squared Residuals")


  spline_fit <- smooth.spline(preds, residuals)
  lines(spline_fit, col = color, lwd = 2)

}


par(mfrow = c(1,1)) 


plot_residuals_base(test_df$logit_preds, test_df$logit_residuals_sq, "Logistic Regression", "red")
plot_residuals_base(test_df$gam_preds, test_df$gam_residuals_sq, "GAM", "blue")
plot_residuals_base(test_df$tree_preds, test_df$tree_residuals_sq, "Decision Tree", "black")



```
### d. 
The tree model's  squared residuals (nearly) flat around 1. 


### e. 

The COMPAS score is an ordinal risk score (1-10), not a probability estimate. 
It does not give a probability of recidivism but rather a ranking of risk.

The squared residuals used in the previous plots rely on a model predicting a probability, 
so this calculation does not work directly.

\newpage


###   Q11.

Based on the evaluation of the logistic regression, GAM, decision tree models, and 
COMPAS, I recommend that Riverdale County adopt the GAM for pre-trial release screening. 
Among the three statistical models, GAM provides the best balance between flexibility 
and interpretability, as its calibration plot shows that predicted probabilities closely 
align with observed recidivism rates,  outperforming the logistic model, which 
exhibited poor calibration at higher probability levels. 

 In terms of classification accuracy and the trade-off 
between false positive and false negative rates, both GAM and logistic regression perform well, 
whereas the decision tree model has more erratic variations across different thresholds. 

Moreover, COMPAS should not be used due to significant limitations: it does not 
output probability estimates, making risk interpretation difficult, and its 
recidivism frequency plot shows irregularities rather than a monotonic increase, 
raising concerns about reliability. 
Given these considerations, GAM is the best choice as it provides a well-calibrated, 
flexible, and interpretable model for Riverdale County’s pre-trial risk assessment. 


\newpage

### Q12.

10 hours



### Bonus.


### Q1. 
### a.

```{r}

library(ggplot2)


test_df$black <- ifelse(test_df$race == "African-American", "Black", "Non-Black")


calculate_fnr_fpr <- function(model_preds, actual) {
  thresholds <- seq(0, 1, length.out=101)
  results <- data.frame(Threshold = thresholds, FNR_Black = NA, FPR_Black = NA, FNR_NonBlack = NA, FPR_NonBlack = NA)

  for (t in thresholds) {
    preds_black <- ifelse(model_preds[test_df$black == "Black"] >= t, 1, 0)
    preds_nonblack <- ifelse(model_preds[test_df$black == "Non-Black"] >= t, 1, 0)

    actual_black <- actual[test_df$black == "Black"]
    actual_nonblack <- actual[test_df$black == "Non-Black"]

    
    FP_Black <- sum(preds_black == 1 & actual_black == 0)
    TN_Black <- sum(preds_black == 0 & actual_black == 0)
    FN_Black <- sum(preds_black == 0 & actual_black == 1)
    TP_Black <- sum(preds_black == 1 & actual_black == 1)

   
    FP_NonBlack <- sum(preds_nonblack == 1 & actual_nonblack == 0)
    TN_NonBlack <- sum(preds_nonblack == 0 & actual_nonblack == 0)
    FN_NonBlack <- sum(preds_nonblack == 0 & actual_nonblack == 1)
    TP_NonBlack <- sum(preds_nonblack == 1 & actual_nonblack == 1)

    
    results$FPR_Black[results$Threshold == t] <- ifelse((FP_Black + TN_Black) > 0, FP_Black / (FP_Black + TN_Black), NA)
    results$FNR_Black[results$Threshold == t] <- ifelse((FN_Black + TP_Black) > 0, FN_Black / (FN_Black + TP_Black), NA)
    
    results$FPR_NonBlack[results$Threshold == t] <- ifelse((FP_NonBlack + TN_NonBlack) > 0, FP_NonBlack / (FP_NonBlack + TN_NonBlack), NA)
    results$FNR_NonBlack[results$Threshold == t] <- ifelse((FN_NonBlack + TP_NonBlack) > 0, FN_NonBlack / (FN_NonBlack + TP_NonBlack), NA)
  }

  return(results)
}

logit_preds <- predict(logit_model, test_df, type="response")
gam_preds <- predict(gam_model, test_df, type="response")
tree_preds <- predict(tree_model, test_df, type="vector")[, 2] 
compas_preds <- test_df$v_score / 10  # Normalize COMPAS scores

logit_results <- calculate_fnr_fpr(logit_preds, test_df$two_year_recid)
gam_results <- calculate_fnr_fpr(gam_preds, test_df$two_year_recid)
tree_results <- calculate_fnr_fpr(tree_preds, test_df$two_year_recid)

plot_fnr_fpr <- function(results, model_name) {
  ggplot(results, aes(x = FPR_Black, y = FNR_Black)) +
    geom_line(color = "blue", size = 1) +
    geom_line(aes(x = FPR_NonBlack, y = FNR_NonBlack), color = "red", size = 1) +
    labs(title = paste("FNR vs. FPR for", model_name),
         x = "False Positive Rate (FPR)",
         y = "False Negative Rate (FNR)",
         color = "Race") +
    scale_color_manual(values = c("Black" = "blue", "Non-Black" = "red")) +
    theme_minimal()
}


plot_fnr_fpr(logit_results, "Logistic Regression")
plot_fnr_fpr(gam_results, "GAM")
plot_fnr_fpr(tree_results, "Decision Tree")


compute_fpr_fnr <- function(t, data) {
  predictions <- ifelse(data$v_score >= t, 1, 0)

  TP <- sum(predictions == 1 & data$two_year_recid == 1)
  FP <- sum(predictions == 1 & data$two_year_recid == 0)
  TN <- sum(predictions == 0 & data$two_year_recid == 0)
  FN <- sum(predictions == 0 & data$two_year_recid == 1)

  FPR <- ifelse((FP + TN) > 0, FP / (FP + TN), NA)
  FNR <- ifelse((FN + TP) > 0, FN / (FN + TP), NA)

  return(c(FPR, FNR))
}

compas_thresholds <- 1:10


error_rates_black <- t(sapply(compas_thresholds, compute_fpr_fnr, data = test_df[test_df$black == "Black", ]))
error_rates_non_black <- t(sapply(compas_thresholds, compute_fpr_fnr, data = test_df[test_df$black == "Non-Black", ]))


fnr_fpr_black <- data.frame(Threshold = compas_thresholds, FPR = error_rates_black[, 1], FNR = error_rates_black[, 2])
fnr_fpr_non_black <- data.frame(Threshold = compas_thresholds, FPR = error_rates_non_black[, 1], FNR = error_rates_non_black[, 2])


ggplot() +
  geom_line(data = fnr_fpr_black, aes(x = FPR, y = FNR, color = "Black"), size = 1) +
  geom_line(data = fnr_fpr_non_black, aes(x = FPR, y = FNR, color = "Non-Black"), size = 1) +
  labs(title = "FNR vs. FPR by Race (COMPAS)",
       x = "False Positive Rate (FPR)",
       y = "False Negative Rate (FNR)",
       color = "Race") +
  scale_color_manual(values = c("Black" = "blue", "Non-Black" = "red")) +
  theme_minimal()

```
### a (continued).

None of the models or COMPAS achieve perfectly equal error rates across racial 
groups at any threshold. However, some models come closer than others.

The decision tree model demonstrates relatively smaller disparities in error rates,
but the curves are still not perfectly aligned.

The GAM model exhibits a more noticeable gap between racial groups in its FNR-FPR 
tradeoff, especially at lower thresholds. 

For COMPAS, the disparity in error rates is visible across different thresholds,
with black individuals exhibiting a higher false negative rate at certain points. 
The logistic regression model shows a similar trend as COMPAS. 


\newpage

### Q2. 

### a.
```{r}

test_df$sex <- as.factor(test_df$sex)
test_df$male <- ifelse(test_df$sex == "Male", "Male", "Female")

compas_scores <- test_df$v_score  
actual_values <- as.numeric(test_df$two_year_recid) 

compas_thresholds <- 1:10 

compute_fpr_fnr <- function(t, data) {
  predictions <- ifelse(data$v_score >= t, 1, 0)  
  
 
  TP <- sum(predictions == 1 & data$two_year_recid == 1)
  FP <- sum(predictions == 1 & data$two_year_recid == 0)
  TN <- sum(predictions == 0 & data$two_year_recid == 0)
  FN <- sum(predictions == 0 & data$two_year_recid == 1)


  FPR <- FP / (FP + TN) 
  FNR <- FN / (FN + TP)  
  
  return(c(FPR, FNR))
}
error_rates_male <- t(sapply(compas_thresholds, compute_fpr_fnr, data = test_df[test_df$male == "Male", ]))
error_rates_female <- t(sapply(compas_thresholds, compute_fpr_fnr, data = test_df[test_df$male == "Female", ]))


fnr_fpr_male <- data.frame(Threshold = compas_thresholds, FPR = error_rates_male[, 1], FNR = error_rates_male[, 2])
fnr_fpr_female <- data.frame(Threshold = compas_thresholds, FPR = error_rates_female[, 1], FNR = error_rates_female[, 2])



ggplot() +
  geom_line(data = fnr_fpr_male, aes(x = FPR, y = FNR, color = "Male"), size = 1) +
  geom_line(data = fnr_fpr_female, aes(x = FPR, y = FNR, color = "Female"), size = 1) +
  labs(title = "FNR vs. FPR by Sex (COMPAS)",
       x = "False Positive Rate (FPR)",
       y = "False Negative Rate (FNR)",
       color = "Sex") +
  theme_minimal()




# logistic model
logit_preds <- predict(logit_model, newdata = test_df, type = "response")

compute_fpr_fnr_model <- function(t, preds, actual) {
  predictions <- ifelse(preds >= t, 1, 0)
  TP <- sum(predictions == 1 & actual == 1)
  FP <- sum(predictions == 1 & actual == 0)
  TN <- sum(predictions == 0 & actual == 0)
  FN <- sum(predictions == 0 & actual == 1)
  
  FPR <- FP / (FP + TN)
  FNR <- FN / (FN + TP)
  return(c(FPR, FNR))
}

logit_error_male <- t(sapply(seq(0.1, 0.9, by = 0.1), compute_fpr_fnr_model, preds = logit_preds[test_df$male == "Male"], actual = actual_values[test_df$male == "Male"]))
logit_error_female <- t(sapply(seq(0.1, 0.9, by = 0.1), compute_fpr_fnr_model, preds = logit_preds[test_df$male == "Female"], actual = actual_values[test_df$male == "Female"]))


logit_fnr_fpr_male <- data.frame(Threshold = seq(0.1, 0.9, by = 0.1), FPR = logit_error_male[, 1], FNR = logit_error_male[, 2])
logit_fnr_fpr_female <- data.frame(Threshold = seq(0.1, 0.9, by = 0.1), FPR = logit_error_female[, 1], FNR = logit_error_female[, 2])


#gam
gam_preds <- predict(gam_model, newdata = test_df, type = "response")

gam_error_male <- t(sapply(seq(0.1, 0.9, by = 0.1), compute_fpr_fnr_model, preds = gam_preds[test_df$male == "Male"], actual = actual_values[test_df$male == "Male"]))
gam_error_female <- t(sapply(seq(0.1, 0.9, by = 0.1), compute_fpr_fnr_model, preds = gam_preds[test_df$male == "Female"], actual = actual_values[test_df$male == "Female"]))

gam_fnr_fpr_male <- data.frame(Threshold = seq(0.1, 0.9, by = 0.1), FPR = gam_error_male[, 1], FNR = gam_error_male[, 2])
gam_fnr_fpr_female <- data.frame(Threshold = seq(0.1, 0.9, by = 0.1), FPR = gam_error_female[, 1], FNR = gam_error_female[, 2])


tree_preds <- predict(tree_model, newdata = test_df, type = "vector")[, 2]

tree_error_male <- t(sapply(seq(0.1, 0.9, by = 0.1), 
                            compute_fpr_fnr_model, preds = tree_preds[test_df$male == "Male"], actual = actual_values[test_df$male == "Male"]))
tree_error_female <- t(sapply(seq(0.1, 0.9, by = 0.1), 
                              compute_fpr_fnr_model, preds = tree_preds[test_df$male == "Female"], actual = actual_values[test_df$male == "Female"]))

tree_fnr_fpr_male <- data.frame(Threshold = seq(0.1, 0.9, by = 0.1), FPR = tree_error_male[, 1], FNR = tree_error_male[, 2])
tree_fnr_fpr_female <- data.frame(Threshold = seq(0.1, 0.9, by = 0.1), FPR = tree_error_female[, 1], FNR = tree_error_female[, 2])

```


```{r}
library(ggplot2)

p1 <- ggplot() +
  geom_line(data = logit_fnr_fpr_male, aes(x = FPR, y = FNR, color = "Male"), size = 1) +
  geom_line(data = logit_fnr_fpr_female, aes(x = FPR, y = FNR, color = "Female"), size = 1) +
  labs(title = "FNR vs. FPR by Sex (Logistic Regression)", 
       x = "False Positive Rate (FPR)", y = "False Negative Rate (FNR)", color = "Sex") +
  theme_minimal()


p2 <- ggplot() +
  geom_line(data = gam_fnr_fpr_male, aes(x = FPR, y = FNR, color = "Male"), size = 1) +
  geom_line(data = gam_fnr_fpr_female, aes(x = FPR, y = FNR, color = "Female"), size = 1) +
  labs(title = "FNR vs. FPR by Sex (GAM Model)", x = "False Positive Rate (FPR)", 
       y = "False Negative Rate (FNR)", color = "Sex") +
  theme_minimal()


p3 <- ggplot() +
  geom_line(data = tree_fnr_fpr_male, aes(x = FPR, y = FNR, color = "Male"), size = 1) +
  geom_line(data = tree_fnr_fpr_female, aes(x = FPR, y = FNR, color = "Female"), size = 1) +
  labs(title = "FNR vs. FPR by Sex (Decision Tree)", x = "False Positive Rate (FPR)", 
       y = "False Negative Rate (FNR)", color = "Sex") +
  theme_minimal()


print(p1)
print(p2)
print(p3)

```
### Bonus 2a(continued).

In every model, there are some differences between the error rates for male and 
female arrestees at most thresholds.
The disparities tend to be larger at lower FPR levels, where female FNR is often 
higher than male FNR.The Decision Tree and GAM models seem to come closest to 
parity, as their lines for male and female arrestees are more aligned compared 
to other models. COMPAS and Logistic Regression show more differentiations. 
